Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-01-07 sources/bmc_template_submission/bmc_article.tex
caption
tighten up discussion

\section*{Discussion}

Databases preserving and democratizing access to biological data
have become essential resources for science.
New molecular data keep accumulating and tools facilitating its integration into
existent evolutionary knowledge contribute to the acceleration of scientific discovery.

To our knowledge, Physcraper is the first tool establishing an interoperability
framework for phylogenetic analyses (should we say this???).
We believe this is a key step to successfully establish an open, reproducible workflow for phylogenetics,
facilitating phylogenetic knowledge for ecologists and other non-specialists, effectively
democratizing phylogenetic studies.

As a tool for automatizing phylogenetic reconstruction from molecular databases,
we believe it presents several advantages over existing phylogenetic pipelines designed to
make evolutionary sense of the vast amount of
public molecular data available. Phylota \cite{sanderson2008phylota},
and PHLAWD \cite{smith2009mega} were the first of the kind and have been cited and used
abundantly for phylogenetic reconstruction and mining of molecular databases. While these are
strong tools that will probably still be widely used with great potential in their
more recent updated versions (phylotar \cite{bennett2018phylotar} and PyPHLAWD \cite{smith2019pyphlawd}),
they do not take into account prior work and knowledge on phylogenetic relationships
and alignment construction, which makes them prone to error (requiring more
manual downstream processing) and difficults comparison with previous phylogenetic
knowledge.

SUPERSMART \cite{antonelli2017toward}) and PUMPER \cite{izquierdo2014pumper}
share conceptual strengths, but present many challenges for installation and user
experience, which have resulted in a low level of adoption of these tools
in the community, and have been applied in very few phylogenetic analyses since their
publication (put number of citations as of today or to harsh??).

Phylogenerator \cite{pearse2013phylogenerator} gathers genetic data by relying only on
genetic marker names and descriptions in the GenBank database, which can be prone
to error due to misslabelling. Any form of sequence similarity analysis, such as BLAST
\cite{camacho2009blast} greatly improves correct homologous identification and should
be preferred for obtaining high-quality phylogenetic hypotheses.

While Physcraper generates individual gene trees, which generally fail to capture the complexity of
species' evolutionary history \cite{song2012resolving}, it facilitates
gathering alignments and gene trees for multiple loci from a group of interest,
that can be used to reconstruct species trees with ASTRAL \cite{mirarab2014astral},
BEAST2 \cite{bouckaert2019beast}, or SVD Quartets \cite{chifman2014quartet}).

Physcraper has the unique advantage to link phylogenies to data available in any of the
taxonomies integrated in the OpenTree taxonomy \cite{rees2017automated}, such as
geographical locations from the Global Biodiversity Information Facility, or
fossils from the Paleobiology Database.
%% a resource that is contributing to achieve the goal of "exploring and
%% analysing biodiversity at an accelerated pace, and returning systematics into
%% the mainstream of science" \cite{wilson2003encyclopedia}.
The Physcraper workflow can be used to rapidly (in a matter of hours)
address challenges overarching both fields of ecology and evolution, such as
phylogenetically placing newly discovered species \cite{webb2010biodiversity},
systematizing molecular (and other) databases, i.e., curating taxonomic assignations
\cite{san2010molecular},
and generating custom trees for ecological \cite{helmus2012phylogenetic} and
evolutionary downstream analyses \cite{stoltzfus2013phylotastic}.

@@

@@

\section*{Figures}
  \begin{figure}[h!]
  \caption{The Physcraper framework consists of 4 general steps (see text). The software is fully
  described on its documentation website at \href{https://physcraper.readthedocs.io}{https://physcraper.readthedocs.io},
  along with installation instructions, function usage descriptions, examples and tutorials.
}
\end{figure}

>>>>

\section*{Discussion}
Databases preserving and democratizing access to biological data
have become essential resources for science.
New molecular data keep accumulating and tools facilitating its integration into
existent evolutionary knowledge contribute to the acceleration of scientific discovery.

We believe this is a key step to successfully establish an open, reproducible workflow for phylogenetics,
facilitating phylogenetic knowledge for ecologists and other non-specialists, effectively
democratizing phylogenetic studies.

As a tool for automatizing phylogenetic reconstruction from molecular databases,
Physcraper presents several advantages over existing phylogenetic pipelines designed to
make evolutionary sense of the vast amount of
public molecular data available.

Several analysis tools create full phylogenies \textit{de novo} by mining of molecular databases \cite{antonelli2017toward, sanderson2008phylota, smith2019pyphlawd, bennett2018phylotar, pearse2013phylogenerator}.
In particular, Phylota \cite{sanderson2008phylota},
and PHLAWD \cite{smith2009mega}, have been cited and used abundantly.

Physcraper builds on this automated database mining concept by incorporating prior phylogenetic work and existing taxonomic domain knowledge on appropriate markers and alignment construction.
This decreases error (requiring more manual downstream processing) and eases comparison with previous phylogenetic
knowledge. 
Unlike placement approaches \cite{berger_performance_2011, matsen_pplacer_2010}, which add new taxa without modifying the input tree, Physcraper estimates all the relationships in the context of the new data.
PUMPER \cite{izquierdo2014pumper} shares these conceptual strengths, but is no longer under active development, is challenging to install and run, and has resulted in very few phylogenetic analyses since its publication.

Physcraper generates gene trees, which individually do not capture the full complexity of species' evolutionary history \cite{song2012resolving}.
However, physcraper facilitates
gathering alignments and gene trees for multiple loci from a group of interest,
that together can be used to reconstruct species trees taking into account coalescent processes with ASTRAL \cite{mirarab2014astral},
BEAST2 \cite{bouckaert2019beast}, or SVD Quartets \cite{chifman2014quartet}).
Rigorous analyses of multiple loci allows for more complex evolutionary models than analyses of large genomic data sets, and can provide better evolutionary estimates.

Physcraper has the added advantage of linking taxonomic information about tips in the output phylogenies to data available in a variety of biological databases \cite{rees2017automated}, such as
geographical locations for taxa from the GBIF \cite{gbif_secretariat_gbif_2019}.

The Physcraper workflow can be used to rapidly (in a matter of hours) to create phylogenies which can address challenges overarching both fields of ecology and evolution, such as
phylogenetically placing newly discovered species \cite{webb2010biodiversity},
curating taxonomic assignments
\cite{san2010molecular},
and generating custom trees for ecological \cite{helmus2012phylogenetic} and
evolutionary downstream analyses \cite{stoltzfus2013phylotastic}.

@@

\textbf{GBIF:} Global Biodiversity Information Facility

@@

\section*{Figures}
  \begin{figure}[h!]
  \caption{The Physcraper framework consists of four general steps. The software is fully
  described on its documentation website at \href{https://physcraper.readthedocs.io}{https://physcraper.readthedocs.io},
  along with installation instructions, function usage descriptions, examples and tutorials.
}
\end{figure}
